---
title: "MuSiC Deconvolution of GSE150910"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

```{r, message=FALSE, include=FALSE}

library(nnls)
library(xbioc)
library(MuSiC)
library(Biobase)
library(reshape2)
library(ggpubr)
library(pheatmap)
library(edgeR)
library(ggrepel)
library(ggplot2)
library(tidyverse)
library(rstatix)
library(enrichR)
library(viridis)
library(factoextra)
library(cowplot)
library(ggvenn)
library(GEOquery)
library(ggExtra)
library(egg)
library(RColorBrewer)
library(openxlsx)

```


```{r, echo=FALSE, message=FALSE}

bulkRaw <- read.csv("/mnt/dzl_bioinf/calverom/Deconvolution/Data/gse150910/GSE150910_gene-level_count_file.csv.gz", header=TRUE, row.names=1)
series <- getGEO(filename="/mnt/dzl_bioinf/calverom/Deconvolution/Data/gse150910/GSE150910_series_matrix.txt.gz", getGPL = FALSE)
metadata <- pData(series)[c(1:2,50:ncol(pData(series)))]
names(metadata)[3:ncol(metadata)] <- c("age", "antigen_identified", "batch", "diagnosis", "ever_smoked", "institution", "immunosuppressant", "lane", "plate", "corticosteroid", "race", "rs35705950", "sample_type", "sex", "tissue")
# remove ipf_H03, has age=NA, voom removes NA
metadata <- droplevels(metadata[metadata$title != "ipf_H03",])
bulkRaw <- bulkRaw[,colnames(bulkRaw) != "ipf_H03"]

metadata$ever_smoked <- ifelse(metadata$ever_smoked == "no", "No", ifelse(metadata$ever_smoked == "yes", "Yes", "Unknown"))
metadata$race[metadata$race == "UNKNOWN"] <- "Unknown"

# remove CHP samples
metadata <- droplevels(metadata[metadata$diagnosis != "chp",])
#metadata2 <- droplevels(metadata2[metadata2$diagnosis != "chp",])
bulkRaw2 <- bulkRaw[,!grepl("^chp", names(bulkRaw))]

metadata$age <- as.numeric(metadata$age)
metadata$diagnosis <- factor(ifelse(metadata$diagnosis == "control", "Control", "IPF"), levels=c("Control", "IPF"))
metadata$sex <- factor(metadata$sex, levels=c("female", "male"))
metadata$ever_smoked <- factor(metadata$ever_smoked, levels=c("No", "Yes", "Unknown"))
metadata <- metadata %>% mutate(race=case_when(race == "1" ~ "Hispanic",
                                               race == "3" ~ "Black",
                                               race == "4" ~ "Asian",
                                               race == "5" ~ "White",
                                               race == "6" ~ "Other",
                                               race == "Unknown" ~ "Unknown"))

rownames(metadata) <- metadata$title
metadata <- metadata[match(colnames(bulkRaw2),rownames(metadata)),]
meta <- data.frame(labelDescription=c("Characters", "Characters", "Numbers", "Factor levels", "Factor levels", "Factor levels", "Factor levels", "Factor levels", "Factor levels", "Factor levels", "Factor levels", "Factor levels", "Factor levels", "Factor levels", "Factor levels", "Factor levels", "Characters"))
phenoData <- new("AnnotatedDataFrame", data=metadata, varMetadata=meta)


# only IPF and control patients in sc data, also downsampled
scKaminski <- readRDS("/mnt/dzl_bioinf/calverom/Deconvolution/Data/Lung_Kaminski_processed_subset_downsampled_new.rds")
#sc_genes <- read_tsv("/mnt/dzl_bioinf/FibrOmics/data/scSEQ/kaminski_GSE136831/features.tsv.gz", col_names=T)


df_sc <- data.frame(SampleID=scKaminski@meta.data$SampleID, Condition=scKaminski@meta.data$Disease, CellType=scKaminski@meta.data$CellType, cluster=scKaminski@meta.data$CellType_Category, row.names = row.names(scKaminski@meta.data))
metaData_sc <- data.frame(labelDescription=c("Factor levels", "Factor levels","Factor levels", "Factor levels"))
phenoData_sc <- new("AnnotatedDataFrame", data=df_sc, varMetadata=metaData_sc)
scKaminski_eset <- ExpressionSet(assayData=as.matrix(scKaminski@assays$RNA@counts), phenoData=phenoData_sc)

```


# QC, normalization of bulk data
```{r}

bulk <- DGEList(counts=bulkRaw2, group=factor(metadata$diagnosis, levels=c("Control", "IPF")))
bulk$genes <- data.frame(GeneName=rownames(bulkRaw2))

# Filter out genes
keep <- filterByExpr(bulk, group=bulk$samples$group)
bulkFiltered <- bulk[keep,,keep.lib.sizes=FALSE]

# Normalization
bulkFiltered <- calcNormFactors(bulkFiltered)
bulkNormLog2cpm <- cpm(bulkFiltered, log=TRUE, normalized.lib.sizes=TRUE)

# add other variables as columns in DGEList object
bulkFiltered$samples$age <- metadata$age
bulkFiltered$samples$race <- metadata$race
bulkFiltered$samples$sex <- metadata$sex
bulkFiltered$samples$ever_smoked <- metadata$ever_smoked
bulkFiltered$samples$plate <- metadata$plate
bulkFiltered$samples$sample <- metadata$sample
bulkFiltered$samples$institution <- metadata$institution
bulkFiltered$samples$batch <- metadata$batch

# bulk as ExpressionSet
bulk_eset <- ExpressionSet(assayData=bulkFiltered$counts, phenoData=phenoData)

```


```{r}

Anno <- data.frame(Disease=metadata$diagnosis, Smoking=metadata$ever_smoked, Sex=metadata$sex)
Anno2 <- data.frame(Disease=metadata$diagnosis, Smoking=metadata$ever_smoked, Sex=metadata$sex, Batch=factor(metadata$batch), Plate=factor(metadata$plate))
rownames(Anno) <- colnames(bulkNormLog2cpm)
rownames(Anno2) <- colnames(bulkNormLog2cpm)
# pheatmap(bulkNormLog2cpm, annotation_col = Anno, annotation_names_col = F, show_rownames = F, show_colnames=F, scale="row")
# pheatmap(bulkNormLog2cpm, annotation_col = Anno2, annotation_names_col = F, show_rownames = F, show_colnames=F, scale="row")

# keep only top 500 most variable genes in heatmap
var_genes <- apply(bulkNormLog2cpm, 1, var)
select_var <- names(sort(var_genes, decreasing=TRUE))[1:500]
mostVar_lcpm <- bulkNormLog2cpm[select_var,]

#jpeg(filename="./supp_figures/heatmap_before.jpg", width=7.25, height=5, units="in", res=350)
pheatmap(mostVar_lcpm, annotation_col = Anno, annotat1.5on_names_col = FALSE, show_rownames = FALSE, show_colnames=FALSE, scale="row", treeheight_row = 0, cellwidth=2, color = colorRampPalette(rev(brewer.pal(n =7, name = "PuOr")))(50)) # without row dendrogram
dev.off()

pheatmap(mostVar_lcpm, annotation_col = Anno2, annotation_names_col = FALSE, show_rownames = FALSE, show_colnames=FALSE, scale="row", treeheight_row = 0, color = colorRampPalette(rev(brewer.pal(n =7, name = "PuOr")))(50))


# similarity between samples
corr <- cor(mostVar_lcpm, method="spearman")
d <- as.dist(1-corr)
pheatmap(d, clustering_distance_rows=d, clustering_distance_cols=d, annotation_col=Anno2, show_colnames=FALSE, show_rownames=FALSE, annotation_names_col=FALSE, color = colorRampPalette(rev(brewer.pal(n =7, name = "PuOr")))(50))
pheatmap(d, clustering_distance_rows=d, clustering_distance_cols=d, annotation_col=Anno, show_colnames=FALSE, show_rownames=FALSE, annotation_names_col=FALSE, color = colorRampPalette(rev(brewer.pal(n =7, name = "PuOr")))(50))

```


# PCA on genes
```{r}

tbulk <- t(mostVar_lcpm)
genePCA <- prcomp(tbulk)
genePCAResSumm <- data.frame(PC=factor(1:nrow(tbulk)), t(summary(genePCA)$importance))
genePCARes <- cbind(metadata, data.frame(genePCA$x))

a <- ggplot(genePCAResSumm[1:20,], aes(x=PC, y=Proportion.of.Variance, group=1)) + geom_point() + geom_line() + theme_classic()
b <- ggplot(genePCARes, aes(x=PC1, y=PC2, color=diagnosis, shape=batch)) + geom_point() + theme_classic()
c <- ggplot(genePCARes, aes(x=PC1, y=PC2, color=diagnosis, shape=plate)) + geom_point() + scale_shape_manual(values=c(19,4,2)) + theme_classic()
d <- ggplot(genePCARes, aes(x=PC1, y=PC2, color=diagnosis, shape=institution)) + geom_point() + theme_classic()

e <- ggplot(genePCARes, aes(x=PC1, y=PC2, color=diagnosis)) + geom_point() + theme_classic()
f <- ggplot(genePCARes, aes(x=PC1, y=PC2, color=sex)) + geom_point() + theme_classic()
g <- ggplot(genePCARes, aes(x=PC1, y=PC2, color=race)) + geom_point() + scale_color_discrete(na.translate=FALSE) + theme_classic()
h <- ggplot(genePCARes, aes(x=PC1, y=PC2, color=ever_smoked)) + geom_point() + scale_color_discrete(na.translate=FALSE) + theme_classic()
i <- ggplot(genePCARes, aes(x=PC1, y=PC2, color=sample_type)) + geom_point() + theme_classic()
j <- ggplot(genePCARes, aes(x=PC1, y=PC2, color=age)) + geom_point() + theme_classic()
plt <- list(a,b,c,d,e,f,g,h,i,j)
# for(pl in 1:length(plt)){
#   ggsave(filename = paste("./supp_figures/scatter_",pl,".jpg",sep=""), plot = plt[[pl]], width=7, height=5, device="jpeg", dpi=350)
# }

```


# Differential expression analysis, adjusting for other variables
```{r}

diagnosis <- factor(bulkFiltered$samples$group, levels=c("Control", "IPF"))
age <- bulkFiltered$samples$age
sex <- factor(bulkFiltered$samples$sex, levels=c("female", "male"))
ever_smoked <- bulkFiltered$samples$ever_smoked
sample <- bulkFiltered$samples$sample
plate <- bulkFiltered$samples$plate
race <- bulkFiltered$samples$race
institution <- bulkFiltered$samples$institution

design <- model.matrix(~0 + diagnosis + age + sex + ever_smoked + sample + plate + race + institution)
colnames(design) <- gsub("diagnosis", "", colnames(design))
colnames(design) <- gsub("sex", "", colnames(design))
colnames(design) <- gsub("sample", "", colnames(design))
colnames(design) <- gsub("institution", "", colnames(design))
contrasts <- makeContrasts(IPF-Control, levels = design)
contrasts
voom <- voom(bulkFiltered, design, plot=TRUE)

vfit <- lmFit(voom, design)
vfit <- contrasts.fit(vfit, contrasts=contrasts)
efit <- eBayes(vfit)
plotSA(efit)

# Number of DE genes, adjusted p-value cutoff = 0.05
edt <- decideTests(efit)
summary(edt)

upreg <- efit$genes$GeneName[which(edt[,1] == 1)]
downreg <- efit$genes$GeneName[which(edt[,1] == -1)] # 8555 total DE genes
res_before_adj <- topTable(efit, coef=1, n=Inf)
res_before_adj

# save de table
#write.xlsx(res_before_adj, file="/mnt/dzl_bioinf/calverom/Deconvolution/music_gse150910/supp_tables/DEtable_before_correction.xlsx", colNames=TRUE, rowNames=FALSE)

DEgenes_before_adj <- c(downreg, upreg)
dfDEgenes_before_adj <- res_before_adj[which(res_before_adj$GeneName %in% DEgenes_before_adj),]

# volcano plot
res_before_adj$updown <- ifelse(res_before_adj$GeneName %in% upreg, "up", ifelse(res_before_adj$GeneName %in% downreg, "down", "notSig"))
options(ggrepel.max.overlaps = Inf)
ggplot(res_before_adj, aes(x=logFC, y=-log10(P.Value), color=updown)) + geom_point(cex=1) + geom_text_repel(data=res_before_adj[1:15,], aes(label=GeneName), size=3) + scale_color_manual(values=c("#0571b0", "grey", "#ca0020"), guide="none") + theme_classic()
#ggsave(filename="./main_figures/volcano_before.jpg", width=7, height=5, device="jpeg", dpi=350)

# Pie chart
freq <- data.frame(summary(edt))
freq <- droplevels(freq[freq$Var1 != "NotSig",])
freq <- freq[order(freq$Var1, decreasing=TRUE),]
freq$prop <- (freq$Freq / sum(freq$Freq))*100
freq$ypos <- cumsum(freq$prop)- 0.5*freq$prop 
ggplot(freq, aes(x="", y=prop, fill=Var1)) + geom_bar(stat="identity", width=1, color="white") + scale_fill_manual(values=c("blue", "red"), labels=c("Down-regulated", "Up-regulated")) + coord_polar("y", start=0) + theme_void() + geom_text(aes(y=ypos, label=Freq), color="white") + labs(fill="")

# DE genes heatmap
#jpeg(filename="./supp_figures/heatmap_before_DEGs.jpg", width=7.25, height=5, units="in", res=350)
pheatmap(bulkNormLog2cpm[DEgenes_before_adj,], annotation_col = Anno, show_rownames=FALSE, show_colnames=FALSE, annotation_names_col=FALSE, scale="row", treeheight_row = 0, cellwidth=2, color = colorRampPalette(rev(brewer.pal(n =7, name = "PuOr")))(50))
dev.off()

```


## Enrichr, adjusted for variables
```{r}

setEnrichrSite("Enrichr")
websiteLive <- TRUE
dbs <- c("BioPlanet_2019", "WikiPathway_2021_Human", "Reactome_2016", "KEGG_2021_Human")

path_before_adj <- enrichr(DEgenes_before_adj, dbs)
plotEnrich(path_before_adj$BioPlanet_2019, showTerms = 10, numChar=60, y = "Ratio", orderBy = "P.value", title="BioPlanet_2019") + scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) + scale_fill_viridis(begin=0.4, option="A") + ggtitle("")
#ggsave(filename="./main_figures/pathways_before.jpg", width=7, height=5, device="jpeg", dpi=350)

plotEnrich(path_before_adj$WikiPathway_2021_Human, showTerms = 10, numChar = 60, y = "Ratio", orderBy = "P.value", title="WikiPathway_2021_Human") + scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) + scale_fill_viridis(begin=0.4, option="A")
plotEnrich(path_before_adj$Reactome_2016, showTerms = 10, numChar=60, y = "Ratio", orderBy = "P.value", title="Reactome_2016") + scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) + scale_fill_viridis(begin=0.4, option="A")
plotEnrich(path_before_adj$KEGG_2021_Human, showTerms = 10, numChar=60, y = "Ratio", orderBy = "Adjusted.P.value", title="KEGG_2021_Human") + scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) + scale_fill_viridis(begin=0.4, option="A")


# save pathway table
#write.xlsx(path_before_adj, file="/mnt/dzl_bioinf/calverom/Deconvolution/music_gse150910/supp_tables/pathwaytable_before_correction.xlsx", colNames=TRUE, rowNames=FALSE)

```


# Deconvolution, using tree
## New clusters
  * Average gene expression per cell type \
  * PCA on the genes, keep top PCs \
  * hclust on new matrix: genes top PCs by cell type
  * keep dendrogram of cell types (column), but original genes on rows in pheatmap
```{r}

# Seurat's AverageExpression() by cell type
ave_exp <- readRDS("/mnt/dzl_bioinf/calverom/Deconvolution/music_ltrc/ave_by_celltype_counts.rds")
gene_pc <- prcomp(t(ave_exp))
summary(gene_pc)
summ <- data.frame(PC=factor(1:ncol(ave_exp)), t(summary(gene_pc)$importance))

ggplot(summ, aes(x=PC, y=Proportion.of.Variance, group=1)) + geom_point() + geom_line() + theme_classic()

```

```{r, fig.height=9, fig.width=9}

# choose top 9 PCs
sub <- gene_pc$x[,1:9] # row = cell types, col = PCs
d <- dist(sub, method = "euclidean")
dend <- hclust(d, method="complete")

plot(dend, hang=-1)
abline(h=10, col="blue", lty=2)

mem10 <- cutree(dend, h=10)
clustersh10 <- data.frame(Cluster = as.numeric(unlist(mem10)), CellType = names(mem10))
clustersh10.list <- split(clustersh10$CellType, clustersh10$Cluster)
names(clustersh10.list) <- paste("C", names(clustersh10.list), sep="")

```

```{r, fig.height=10, fig.width=11}

pheatmap(ave_exp, cluster_cols=dend, scale="row", show_rownames=FALSE, treeheight_col = 150, cutree_cols = 20, color = colorRampPalette(rev(brewer.pal(n =7, name = "PuOr")))(50))

```

```{r}

# use new clusters 
new.markers <- readRDS("/mnt/dzl_bioinf/calverom/Deconvolution/music_ltrc/markers_by_newestclusters_Kaminski.rds")
new.markers.list <- NULL
for(cl in unique(new.markers$cluster)){
  markrs <- new.markers$gene[new.markers$cluster == cl & new.markers$p_val_adj <= 0.05]
  new.markers.list[[cl]] <- markrs
}

clusters <- c("Lymphatic", "VE_Venous", "VE_Arterial", "cMonocyte", "ncMonocyte", "Fibroblast", "Myofibroblast", "Basal", "Club", "Goblet", "Aberrant_Basaloid", "cDC2", "DC_Langerhans", "NK", "ILC_A", "T", "T_Cytotoxic", "ILC_B", "Mast", "ATI", "VE_Capillary_A", "VE_Capillary_B", "Macrophage_Alveolar", "Macrophage", "B_Plasma", "B", "T_Regulatory", "Mesothelial", "Ciliated", "ATII", "cDC1", "DC_Mature", "pDC", "SMC", "Pericyte", "VE_Peribronchial", "Ionocyte", "PNEC")
names(clusters) <- c(rep("C1", 3), rep("C2", 2), rep("C3", 2), rep("C4", 4), rep("C5", 2), rep("C6", 6), "C7", rep("C8", 2), rep("C9", 2), "C10", rep("C11",2), "C12", "C13", "C14", rep("C15", 2), "C16", rep("C17",2), "C18", "C19", "C20")
scKaminski_eset$new_clusters <- as.character(names(clusters)[match(scKaminski_eset$CellType, clusters)])

a <- Sys.time()
musicRes_new <- music_prop.cluster(bulk.eset=bulk_eset, sc.eset=scKaminski_eset, group.markers=new.markers.list, clusters="CellType", groups="new_clusters", samples="SampleID", clusters.type=clustersh10.list, verbose=FALSE)
b <- Sys.time()

celltypes <- unique(scKaminski_eset$CellType)


# save props table
#write.xlsx(musicRes_new$Est.prop.weighted, file="/mnt/dzl_bioinf/calverom/Deconvolution/music_gse150910/supp_tables/deconvolution_results.xlsx", rowNames=FALSE, colNames=TRUE)

```

```{r}

dfEst <- data.frame(musicRes_new$Est.prop.weighted, Diagnosis=factor(pData(bulk_eset)$diagnosis, levels=c("Control", "IPF")), Age=pData(bulk_eset)$age, Race=factor(pData(bulk_eset)$race), Sex=factor(pData(bulk_eset)$sex), Ever_smoked=factor(pData(bulk_eset)$ever_smoked), Sample=factor(pData(bulk_eset)$sample), Plate=factor(pData(bulk_eset)$plate), Institution=factor(pData(bulk_eset)$institution), check.names=FALSE)


# sorted by Control proportions
melted <- melt(dfEst, id.vars=c("Diagnosis", "Age", "Race", "Sex", "Ever_smoked", "Sample", "Plate", "Institution"))
ord <- data.frame(dfEst %>% group_by(Diagnosis) %>% summarize_at(vars(Lymphatic:PNEC), median), check.names=FALSE)
ord <- melt(ord)
ord <- ord[order(ord$Diagnosis, -ord$value),]
melted$variable <- factor(melted$variable, levels=unique(ord$variable))

# no significance stars
ggplot(melted, aes(x=variable, y=value)) + geom_boxplot(aes(color=Diagnosis), position=position_dodge(1), outlier.size = 0.5) + scale_color_manual(values=rev(c("#B2182B", "#2166AC")),  guide = guide_legend(title = "")) + xlab("") + ylab("Estimated proportion") + theme_classic() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), legend.position=c(0.95,0.9))


# Not including cell types with all zero proportions
colSum <- colSums(dfEst[,1:38])
nonzeroCT <- names(colSum)[colSum != 0]
tmp <- dfEst[,names(dfEst) %in% c(nonzeroCT, "Diagnosis")]

# sort by Control proportions
melted <- melt(tmp)
ord <- data.frame(tmp %>% group_by(Diagnosis) %>% summarize_all("median"), check.names=FALSE)
ord <- melt(ord)
ord <- ord[order(ord$Diagnosis, -ord$value),]
names(melted)[2] <- "celltype"
melted$celltype <- factor(melted$celltype, levels=unique(ord$variable))

# without significance stars, p values as bars
stat.test <- melted %>% group_by(celltype) %>% wilcox_test(value ~ Diagnosis, exact=FALSE) %>% adjust_pvalue(method = "BH") %>% add_significance(p.col="p") %>% add_significance(p.col="p.adj")

a <- ggplot(melted, aes(x=celltype, y=value)) + geom_boxplot(aes(color=Diagnosis), position=position_dodge(1), outlier.size = 0.5) + scale_color_manual(values=rev(c("#B2182B", "#2166AC")),  guide = guide_legend(title = "")) + xlab("") + ylab("Estimated proportion") + theme_classic() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), legend.position=c(0.95,0.9), plot.margin = unit(c(0.1,0.1,-0.95,0.1), "cm"))

b <- ggplot(stat.test, aes(x=celltype, y=-log10(p))) + geom_bar(stat="identity", width=0.6) + scale_y_reverse(expand = c(0, 0.05) , limits=c(19,0)) + scale_x_discrete(position = "top") + xlab("") + theme_classic() + theme(axis.text.x = element_blank(), panel.background = element_rect(color = "black"))
egg::ggarrange(a,b, nrow=2, heights=c(3,1))


# with significance stars
stat.test <- melted %>%
  group_by(celltype) %>%
  wilcox_test(value ~ Diagnosis, exact=FALSE) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance(p.col="p") %>% add_significance(p.col="p.adj")

# boxplot
bxp <- ggplot(melted, aes(x=celltype, y=value)) + geom_boxplot(aes(color=Diagnosis), position=position_dodge(0.8), outlier.size = 0.5, width=0.65) + scale_color_manual(values=rev(c("#B2182B", "#2166AC")),  guide = guide_legend(title = "")) + xlab("") + ylab("Estimated proportion") + theme_classic() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), legend.position=c(0.95,0.9))
stat.test <- stat.test %>% filter(p.adj.signif != "ns") %>% add_xy_position(x="celltype", dodge=0.8)
stat.test$y.position2 <- c(max(dfEst$ATI) + 0.02, max(dfEst$VE_Capillary_B) + 0.02, max(dfEst$VE_Capillary_A) + 0.02, max(dfEst$Ciliated) + 0.02, max(dfEst$Fibroblast) + 0.02, max(dfEst$Myofibroblast) + 0.02, max(dfEst$Lymphatic) + 0.02, max(dfEst$cMonocyte) + 0.02, max(dfEst$VE_Venous) + 0.02, max(dfEst$ncMonocyte) + 0.02, max(dfEst$VE_Arterial) + 0.02, max(dfEst$cDC2) + 0.02, max(dfEst$DC_Langerhans) + 0.02, max(dfEst$T) + 0.02, max(dfEst$B_Plasma) + 0.02, max(dfEst$pDC) + 0.02, max(dfEst$Ionocyte) + 0.02, max(dfEst$PNEC) + 0.02)


bxp + stat_pvalue_manual(stat.test, y.position="y.position2", label = "p.adj.signif", tip.length = 0.005, size=2.5)
#ggsave(filename="./main_figures/deconvolution.jpg", width=9, height=7, device="jpeg", dpi=350)

```


# Test difference of proportions between Control and IPF
```{r, warning=FALSE}

wilcoxRes <- NULL
for(cell in names(dfEst)[1:38]){
  wilcox <- wilcox.test(dfEst[,cell] ~ dfEst[,"Diagnosis"], data=dfEst)
  wilcoxRes <- rbind(wilcoxRes, data.frame(cellType=cell, W=wilcox$statistic, pVal=wilcox$p.value))
}

rownames(wilcoxRes) <- NULL
hist(wilcoxRes$pVal, breaks=100)
wilcoxRes$pAdj <- p.adjust(wilcoxRes$pVal, method="BH")
wilcoxRes[order(wilcoxRes$pAdj),]

none <- wilcoxRes$cellType[is.nan(wilcoxRes$pVal)]
orderCT <- wilcoxRes$cellType[order(wilcoxRes$pAdj)]
orderCT <- orderCT[!(orderCT %in% none)]

```


# Correlation, all genes (with variable corrections) vs estimated cell proportions
```{r}

corrRes_CellPropxallGene2 <- NULL
for(ct in setdiff(celltypes,none)){
  corr <- apply(bulkNormLog2cpm, 1, cor.test, dfEst[,ct], method="pearson")
  genes <- names(corr)
  corrRes_CellPropxallGene2 <- rbind(corrRes_CellPropxallGene2, data.frame(cellType=ct, gene=genes, matrix(unlist(corr), nrow=length(corr), byrow=TRUE)))
}

corrRes_CellPropxallGene2 <- corrRes_CellPropxallGene2[,c(1:2,6,3,5)]
colnames(corrRes_CellPropxallGene2)[3:5] <- c("corrCoefficient", "testStatistic", "pVal")
corrRes_CellPropxallGene2[,3:5] <- lapply(corrRes_CellPropxallGene2[,3:5], as.numeric)
corrRes_CellPropxallGene2$pValDEA <- res_before_adj$P.Value[match(corrRes_CellPropxallGene2$gene, res_before_adj$GeneName)]
corrRes_CellPropxallGene2$pAdj <- p.adjust(corrRes_CellPropxallGene2$pVal, method="BH")
corrRes_CellPropxallGene2[order(corrRes_CellPropxallGene2$pAdj),]
corrRes_CellPropxallGene2$updown <- res_before_adj$updown[match(corrRes_CellPropxallGene2$gene, res_before_adj$GeneName)]

for(ct in orderCT){
  print(ggplot(corrRes_CellPropxallGene2[which(corrRes_CellPropxallGene2$cellType == ct),], aes(x=corrCoefficient, y=-log10(pValDEA))) + geom_point(aes(color=updown), alpha=0.4) + scale_color_manual(name="up/down reg", values=c("blue", "black", "red")) + xlab("correlation coefficient") + ylab("-log10(DEA p-value)") + labs(title=ct) + theme_classic())
}

```


# PCA on estimated cell proportions
```{r}

cellProps <- dfEst[,which(!(names(dfEst)[1:38] %in% none))]
cellPCA <- prcomp(cellProps)
cellPCAResSumm <- data.frame(PC=factor(1:ncol(cellProps)), t(summary(cellPCA)$importance))
cellPCARes <- cbind(dfEst, data.frame(cellPCA$x))
cellLoadings <- data.frame(cellPCA$rotation)


a <- ggplot(cellPCAResSumm[1:20,], aes(x=PC, y=Proportion.of.Variance, group=1)) + geom_point() + geom_line() + theme_classic()
b <- ggplot(cellPCAResSumm[1:20,], aes(x=PC, y=Cumulative.Proportion, group=1)) + geom_point() + geom_line() + theme_classic()


# biplot
ggplot(cellLoadings, aes(x=PC1, y=PC2, label=rownames(cellLoadings))) + geom_point(cex=1) + geom_text_repel(size=3.5)

fviz_pca_biplot(cellPCA, axes=c(1,2), geom.ind="point", col.ind=cellPCARes$Diagnosis, select.var=list(name=c("Ciliated", "Macrophage", "VE_Capillary_B", "VE_Capillary_A", "Macrophage_Alveolar", "ATI", "cMonocyte", "Fibroblast")), invisible="quali", legend.title="Diagnosis", repel=TRUE) + scale_shape_manual(values=c(19,19)) + scale_color_manual(values=rev(c("#F8766D", "#00BFC4"))) + theme_minimal()
#ggsave("./main_figures/PCA_dim12.jpg", width=7, height=5, device="jpeg", dpi=350)

c <- fviz_pca_biplot(cellPCA, axes=c(1,3), geom.ind="point", col.ind=cellPCARes$Diagnosis, select.var=list(name=c("Ciliated", "Macrophage", "VE_Capillary_B", "VE_Capillary_A", "Macrophage_Alveolar", "ATI", "cMonocyte", "Fibroblast")), invisible="quali", legend.title="Diagnosis", repel=TRUE) + scale_shape_manual(values=c(19,19)) + scale_color_manual(values=rev(c("#F8766D", "#00BFC4"))) + theme_minimal()

d <- fviz_pca_biplot(cellPCA, axes=c(2,3), geom.ind="point", col.ind=cellPCARes$Diagnosis, select.var=list(name=c("Ciliated", "Macrophage", "VE_Capillary_B", "VE_Capillary_A", "Macrophage_Alveolar", "ATI", "cMonocyte", "Fibroblast")), invisible="quali", legend.title="Diagnosis", repel=TRUE) + scale_shape_manual(values=c(19,19)) + scale_color_manual(values=rev(c("#F8766D", "#00BFC4"))) + theme_minimal()

pl <- list(a,b,c,d)
# for(i in 1:length(pl)){
#   ggsave(filename = paste("./supp_figures/PCA_",i,".jpg",sep=""), plot = pl[[i]], width=7, height=5, device="jpeg", dpi=350)
# }


```


# Differential expression analysis (adjusting for cell proportions), gene ~ diagnosis + age + sex + smoking_history + sample + plate + race + institution + PC1 cell proportions
```{r}

bulkFiltered_add <- bulkFiltered
bulkFiltered_add$samples$PC1 <- cellPCARes[,"PC1"]

diagnosis <- bulkFiltered_add$samples$group
age <- bulkFiltered_add$samples$age
sex <- factor(bulkFiltered_add$samples$sex, levels=c("female", "male"))
ever_smoked <- bulkFiltered_add$samples$ever_smoked
sample <- bulkFiltered_add$samples$sample
plate <- bulkFiltered_add$samples$plate
race <- bulkFiltered_add$samples$race
institution <- bulkFiltered_add$samples$institution

PC1 <- cellPCARes[,"PC1"]

design2 <- model.matrix(~0 + diagnosis + age + sex + ever_smoked + sample + plate + race + institution + PC1)
colnames(design2) <- gsub("diagnosis", "", colnames(design2))
colnames(design2) <- gsub("sex", "", colnames(design2))
colnames(design2) <- gsub("sample", "", colnames(design2))
colnames(design2) <- gsub("institution", "", colnames(design2))
contrasts2 <- makeContrasts(IPF-Control, levels = design2)
contrasts2
voom2 <- voom(bulkFiltered_add, design2, plot=TRUE)

vfit2 <- lmFit(voom2, design2)
vfit2 <- contrasts.fit(vfit2, contrasts=contrasts2)
efit2 <- eBayes(vfit2)
plotSA(efit2)


# Number of DE genes, adjusted p-value cutoff = 0.05
edt2 <- decideTests(efit2)
summary(edt2)

upreg <- efit2$genes$GeneName[which(edt2[,1] == 1)]
downreg <- efit2$genes$GeneName[which(edt2[,1] == -1)]
res_afterPC1 <- topTable(efit2, coef=1, n=Inf)
res_afterPC1

# save de table
#write.xlsx(res_afterPC1, file="/mnt/dzl_bioinf/calverom/Deconvolution/music_gse150910/supp_tables/DEtable_after_PC1correction.xlsx", rowNames=FALSE, colNames=TRUE)

DEgenes_afterPC1 <- c(downreg, upreg)
dfDEgenes_afterPC1 <- res_afterPC1[which(res_afterPC1$GeneName %in% DEgenes_afterPC1),]

# volcano plot
res_afterPC1$updown <- ifelse(res_afterPC1$GeneName %in% upreg, "up", ifelse(res_afterPC1$GeneName %in% downreg, "down", "notSig"))
options(ggrepel.max.overlaps = Inf)
ggplot(res_afterPC1, aes(x=logFC, y=-log10(P.Value), color=updown)) + geom_point(cex=1) + geom_text_repel(data=res_afterPC1[1:15,], aes(label=GeneName), size=3) + scale_color_manual(values=c("#0571b0", "grey", "#ca0020"), guide="none") + theme_classic()
#ggsave(filename = "./supp_figures/volcano_afterPC1.jpg", width=7, height=5, device="jpeg", dpi=350)


# Pie chart
freq <- data.frame(summary(edt2))
freq <- droplevels(freq[freq$Var1 != "NotSig",])
freq <- freq[order(freq$Var1, decreasing=TRUE),]
freq$prop <- (freq$Freq / sum(freq$Freq))*100
freq$ypos <- cumsum(freq$prop)- 0.5*freq$prop 
ggplot(freq, aes(x="", y=prop, fill=Var1)) + geom_bar(stat="identity", width=1, color="white") + scale_fill_manual(values=c("blue", "red"), labels=c("Down-regulated", "Up-regulated")) + coord_polar("y", start=0) + theme_void() + geom_text(aes(y=ypos, label=Freq), color="white") + labs(fill="")

# DE genes heatmap
#jpeg(filename="./supp_figures/heatmap_afterPC1_DEGs.jpg", width=7.25, height=5, units="in", res=350)
pheatmap(bulkNormLog2cpm[DEgenes_afterPC1,], annotation_col = Anno, show_rownames=FALSE, show_colnames=FALSE, annotation_names_col=FALSE, scale="row", color = colorRampPalette(rev(brewer.pal(n =7, name = "PuOr")))(50), treeheight_row = 0, cellwidth=2)
dev.off()

```


```{r}

# Venn diagram
ggvenn(list(`with variables correction`=DEgenes_before_adj,  `with PC1 correction`=DEgenes_afterPC1), show_percentage = FALSE, set_name_size = 3, fill_color=c("#56B4E9", "#F0E442"))

```


# Enrichr, PC1 correction
```{r}

path_afterPC1 <- enrichr(DEgenes_afterPC1, dbs)

plotEnrich(path_afterPC1$BioPlanet_2019, showTerms = 10, numChar=60, y = "Ratio", orderBy = "P.value", title="BioPlanet_2019") + scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) + scale_fill_viridis(begin=0.4, option="A") + ggtitle("")
#ggsave("./main_figures/pathways_afterPC1.jpg", width=7, height=5, device="jpeg", dpi=350)

plotEnrich(path_afterPC1$WikiPathway_2021_Human, showTerms = 10, numChar = 60, y = "Ratio", orderBy = "P.value", title="WikiPathway_2021_Human") + scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) + scale_fill_viridis(begin=0.4, option="A")
plotEnrich(path_afterPC1$Reactome_2016, showTerms = 10, numChar=60, y = "Ratio", orderBy = "P.value", title="Reactome_2016") + scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) + scale_fill_viridis(begin=0.4, option="A")
plotEnrich(path_afterPC1$KEGG_2021_Human, showTerms = 10, numChar=60, y = "Ratio", orderBy = "Adjusted.P.value", title="KEGG_2021_Human") + scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) + scale_fill_viridis(begin=0.4, option="A")

# save pathway table
#write.xlsx(path_afterPC1, file="/mnt/dzl_bioinf/calverom/Deconvolution/music_gse150910/supp_tables/pathwaytable_after_PC1correction.xlsx", rowNames=FALSE, colNames=TRUE)

```


# Correlation, all genes vs estimated cell proportions, PC1 correction
```{r}

corrRes_CellPropxallGene_afterPC1 <- NULL

for(ct in setdiff(celltypes,none)){
  corr <- apply(bulkNormLog2cpm, 1, cor.test, dfEst[,ct], method="pearson")
  genes <- names(corr)
  corrRes_CellPropxallGene_afterPC1 <- rbind(corrRes_CellPropxallGene_afterPC1, data.frame(cellType=ct, gene=genes, matrix(unlist(corr), nrow=length(corr), byrow=TRUE)))
}

corrRes_CellPropxallGene_afterPC1 <- corrRes_CellPropxallGene_afterPC1[,c(1:2,6,3,5)]
colnames(corrRes_CellPropxallGene_afterPC1)[3:5] <- c("corrCoefficient", "testStatistic", "pVal")
corrRes_CellPropxallGene_afterPC1[,3:5] <- lapply(corrRes_CellPropxallGene_afterPC1[,3:5], as.numeric)
corrRes_CellPropxallGene_afterPC1$pValDEA <- res_afterPC1$P.Value[match(corrRes_CellPropxallGene_afterPC1$gene, res_afterPC1$GeneName)]
corrRes_CellPropxallGene_afterPC1$pAdj <- p.adjust(corrRes_CellPropxallGene_afterPC1$pVal, method="BH")
corrRes_CellPropxallGene_afterPC1[order(corrRes_CellPropxallGene_afterPC1$pAdj),]
corrRes_CellPropxallGene_afterPC1$updown <- res_afterPC1$updown[match(corrRes_CellPropxallGene_afterPC1$gene, res_afterPC1$GeneName)]
# save corr table
#write.xlsx(corrRes_CellPropxallGene_afterPC1, "/mnt/dzl_bioinf/calverom/Deconvolution/music_gse150910/supp_tables/correlation_celltype_genes_afterPC1_correction.xlsx", rowNames=FALSE, colNames=TRUE)

for(ct in orderCT){
  print(ggplot(corrRes_CellPropxallGene_afterPC1[which(corrRes_CellPropxallGene_afterPC1$cellType == ct),], aes(x=corrCoefficient, y=-log10(pValDEA))) + geom_point(aes(color=updown), alpha=0.4) + scale_color_manual(name="up/down reg", values=c("blue", "black", "red")) + xlab("correlation coefficient") + ylab("-log10(DEA p-value)") + annotate("text", x=-Inf, y=Inf, label=ct, hjust=-0.1, vjust=1, size=5) + theme_classic() + theme(legend.position="top"))
  pathh <- paste("./main_figures/ears_PC1/ears_afterPC1", "_", ct, ".jpg", sep="")
  #ggsave(filename=pathh, width=7, height=5, device="jpeg", dpi=350)
}

```

## Enrichr, cell type-specific pathways in bulk, PC1 correction
```{r}

# For EnrichR

# subset with correlation coefficient > 0.5
corrRes_afterPC1_sub <- corrRes_CellPropxallGene_afterPC1[corrRes_CellPropxallGene_afterPC1$corrCoefficient >= 0.5,]
corrRes_afterPC1_sub <- corrRes_afterPC1_sub[,-c(6,8)]
#corrRes_afterPC1_sub$gene2 <- geneAnno$GeneName[match(corrRes_afterPC1_sub$gene, geneAnno$ensembl)]
corrGenes <- unique(corrRes_afterPC1_sub$gene)
#corrGenes_conv <- geneAnno$GeneName[match(corrGenes, geneAnno$ensembl)]

# subset DE results with only genes from above^
res_afterPC1_sub <- res_afterPC1[res_afterPC1$GeneName %in% corrGenes,]
res_afterPC1_sub <- res_afterPC1_sub[,1:ncol(res_afterPC1_sub)-1]
# recompute adj.P.Val
res_afterPC1_sub$adj.P.Val <- p.adjust(res_afterPC1_sub$P.Value, method="BH")

# add p-values and adjusted p-values from above^ to corrRes_afterPC1_sub
corrRes_afterPC1_sub$pValDEA <- res_afterPC1_sub$P.Value[match(corrRes_afterPC1_sub$gene, res_afterPC1_sub$GeneName)]
corrRes_afterPC1_sub$pAdjDEA <- res_afterPC1_sub$adj.P.Val[match(corrRes_afterPC1_sub$gene, res_afterPC1_sub$GeneName)]

corrRes_afterPC1_sub$sig0.05 <- ifelse(corrRes_afterPC1_sub$pAdjDEA <= 0.05, "sig", "not sig")
corrRes_afterPC1_sub$sig0.1 <- ifelse(corrRes_afterPC1_sub$pAdjDEA <= 0.1, "sig", "not sig")
# count DE genes (alpha = 0.05) per cell type
s0.05 <- as.data.frame(table(corrRes_afterPC1_sub$cellType, corrRes_afterPC1_sub$sig0.05))
# count DE genes (alpha=0.1) per cell type
s0.1 <- as.data.frame(table(corrRes_afterPC1_sub$cellType, corrRes_afterPC1_sub$sig0.1))

# write.xlsx(corrRes_afterPC1_sub, "./results tree 22.4.25/afterPC1_correction_corr_CTxsubGenes.xlsx")

ggplot(s0.05[s0.05$Var2 == "sig",], aes(x=Var1, y=Freq, fill=Var2)) + geom_bar(stat="identity", position="dodge") + xlab("") + ylab("number of genes (for Enrichr)") + theme_classic() + coord_flip() + theme(legend.position="none")

```

```{r, out.extra='style="max-width:none; width:80vw; margin-left:calc(50% - 40vw);"', fig.width = 17, fig.height=10}

path_afterPC1_celltype_specific <- NULL
for(ct in orderCT){
  temp <- droplevels(corrRes_afterPC1_sub[corrRes_afterPC1_sub$cellType == ct, ])
  genes <- temp$gene[temp$sig0.05 == "sig"]
  if(length(genes) >= 10){
    enr <- enrichr(genes, dbs)
    path_afterPC1_celltype_specific[[ct]] <- enr
    p1 <- plotEnrich(path_afterPC1_celltype_specific[[ct]]$BioPlanet_2019, showTerms = 10, numChar=60, y = "Ratio", orderBy = "P.value", title="BioPlanet_2019") + scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) + scale_fill_viridis(begin=0.4, option="A") + ggtitle("")
    pathh <- paste("./main_figures/pathways_celltype_PC1/path_afterPC1", "_", ct, ".jpg", sep="")
    #ggsave(filename=pathh, plot=p1, width=7, height=5, device="jpeg", dpi=350)
    
    p2 <- plotEnrich(path_afterPC1_celltype_specific[[ct]]$WikiPathway_2021_Human, showTerms = 10, numChar=60, y = "Ratio", orderBy = "P.value", title="WikiPathway_2021_Human") + scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) + scale_fill_viridis(begin=0.4, option="A")
    p3 <- plotEnrich(path_afterPC1_celltype_specific[[ct]]$Reactome_2016, showTerms = 10, numChar=60, y = "Ratio", orderBy = "P.value", title="Reactome_2016") + scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) + scale_fill_viridis(begin=0.4, option="A")
    title <- ggdraw() + draw_label(ct, x=0, hjust=0, size=12) + theme(plot.margin = margin(0, 0, 0, 7))
    pl <- plot_grid(p1,p2,p3, ncol=2)
    print(plot_grid(title,pl, ncol=1,rel_heights=c(0.1, 1)))
  }
}

# write to excel, pathway results per cell type
#write_rds(path_afterPC1_celltype_specific, "/mnt/dzl_bioinf/calverom/Deconvolution/music_gse150910/CTpathways_PC1corrected.rds")
#write_rds(path_afterPC1_celltype_specific, "/mnt/dzl_bioinf/calverom/Deconvolution/music_gse150910/supp_tables/CTpathways_PC1corrected.rds")


```


# Differential expression analysis (adjusting for cell proportions), gene ~ diagnosis + age + sex + smoking_history + sample + plate + race + institution + PC1 + PC2 + PC3 + PC4 cell proportions
```{r}

bulkFiltered_add <- bulkFiltered
bulkFiltered_add$samples$PC1 <- cellPCARes[,"PC1"]
bulkFiltered_add$samples$PC2 <- cellPCARes[,"PC2"]
bulkFiltered_add$samples$PC3 <- cellPCARes[,"PC3"]
bulkFiltered_add$samples$PC4 <- cellPCARes[,"PC4"]

diagnosis <- bulkFiltered_add$samples$group
age <- bulkFiltered_add$samples$age
sex <- factor(bulkFiltered_add$samples$sex, levels=c("female", "male"))
ever_smoked <- bulkFiltered_add$samples$ever_smoked
sample <- bulkFiltered_add$samples$sample
plate <- bulkFiltered_add$samples$plate
race <- bulkFiltered_add$samples$race
institution <- bulkFiltered_add$samples$institution

PC1 <- cellPCARes[,"PC1"]
PC2 <- cellPCARes[,"PC2"]
PC3 <- cellPCARes[,"PC3"]
PC4 <- cellPCARes[,"PC4"]

design3 <- model.matrix(~0 + diagnosis + age + sex + ever_smoked + sample + plate + race + institution + PC1 + PC2 + PC3 + PC4)
colnames(design3) <- gsub("diagnosis", "", colnames(design3))
colnames(design3) <- gsub("sex", "", colnames(design3))
colnames(design3) <- gsub("sample", "", colnames(design3))
colnames(design3) <- gsub("institution", "", colnames(design3))
contrasts3 <- makeContrasts(IPF-Control, levels = design3)
contrasts3
voom3 <- voom(bulkFiltered_add, design3, plot=TRUE)

vfit3 <- lmFit(voom3, design3)
vfit3 <- contrasts.fit(vfit3, contrasts=contrasts3)
efit3 <- eBayes(vfit3)
plotSA(efit3)


# Number of DE genes, adjusted p-value cutoff = 0.05
edt3 <- decideTests(efit3)
summary(edt3)

upreg <- efit3$genes$GeneName[which(edt3[,1] == 1)]
downreg <- efit3$genes$GeneName[which(edt3[,1] == -1)]
res_afterPC1_PC4 <- topTable(efit3, coef=1, n=Inf)
res_afterPC1_PC4

#save de table
#write.xlsx(res_afterPC1_PC4, "/mnt/dzl_bioinf/calverom/Deconvolution/musicgse150910/supp_tables/DEtable_after_PC1_PC4correction.xlsx", rowNames=FALSE, colNames=TRUE)

DEgenes_afterPC1_PC4 <- c(downreg, upreg)
dfDEgenes_afterPC1_PC4 <- res_afterPC1_PC4[which(res_afterPC1_PC4$GeneName %in% DEgenes_afterPC1_PC4),]

# volcano plot
res_afterPC1_PC4$updown <- ifelse(res_afterPC1_PC4$GeneName %in% upreg, "up", ifelse(res_afterPC1_PC4$GeneName %in% downreg, "down", "notSig"))
options(ggrepel.max.overlaps = Inf)

ggplot(res_afterPC1_PC4, aes(x=logFC, y=-log10(P.Value), color=updown)) + geom_point(cex=1) + geom_text_repel(data=res_afterPC1_PC4[1:15,], aes(label=GeneName), size=3) + scale_color_manual(values=c("#0571b0", "grey", "#ca0020"), guide="none") + theme_classic()
#ggsave(filename="./supp_figures/volcano_afterPC1PC4.jpg", width=7, height=5, dpi=350)

# Pie chart
freq <- data.frame(summary(edt3))
freq <- droplevels(freq[freq$Var1 != "NotSig",])
freq <- freq[order(freq$Var1, decreasing=TRUE),]
freq$prop <- (freq$Freq / sum(freq$Freq))*100
freq$ypos <- cumsum(freq$prop)- 0.5*freq$prop 
ggplot(freq, aes(x="", y=prop, fill=Var1)) + geom_bar(stat="identity", width=1, color="white") + scale_fill_manual(values=c("blue", "red"), labels=c("Down-regulated", "Up-regulated")) + coord_polar("y", start=0) + theme_void() + geom_text(aes(y=ypos, label=Freq), color="white") + labs(fill="")

# DE genes heatmap
#jpeg(filename="./supp_figures/heatmap_afterPC1PC4_DEGs.jpg", width=7.25, height=5, units="in", res=350)
pheatmap(bulkNormLog2cpm[DEgenes_afterPC1_PC4,], annotation_col = Anno, show_rownames=FALSE, show_colnames=FALSE, annotation_names_col=FALSE, scale="row", color = colorRampPalette(rev(brewer.pal(n =7, name = "PuOr")))(50), treeheight_row = 0, cellwidth=2)
dev.off()

```


```{r}

# Venn diagram
ggvenn(list(`with variables correction`=DEgenes_before_adj,  `with PC1-PC4 correction`=DEgenes_afterPC1_PC4), show_percentage = FALSE, set_name_size = 3, fill_color=c("#56B4E9", "#F0E442"))

ggvenn(list(`with variables correction`=DEgenes_before_adj, `with PC1 correction`=DEgenes_afterPC1,  `with PC1-PC4 correction`=DEgenes_afterPC1_PC4), show_percentage = FALSE, set_name_size = 3, fill_color=c("#56B4E9", "#F0E442", "#CC79A7"))

```


# Enrichr, PC1-PC4 correction
```{r}

path_afterPC1_PC4 <- enrichr(DEgenes_afterPC1_PC4, dbs)

plotEnrich(path_afterPC1_PC4$BioPlanet_2019, showTerms = 10, numChar=60, y = "Ratio", orderBy = "P.value", title="BioPlanet_2019") + scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) + scale_fill_viridis(begin=0.4, option="A") + ggtitle("")
#ggsave("./main_figures/pathways_afterPC1PC4.jpg", width=7, height=5, device="jpeg", dpi=350)

plotEnrich(path_afterPC1_PC4$WikiPathway_2021_Human, showTerms = 10, numChar = 60, y = "Ratio", orderBy = "P.value", title="WikiPathway_2021_Human") + scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) + scale_fill_viridis(begin=0.4, option="A")
plotEnrich(path_afterPC1_PC4$Reactome_2016, showTerms = 10, numChar=60, y = "Ratio", orderBy = "P.value", title="Reactome_2016") + scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) + scale_fill_viridis(begin=0.4, option="A")
plotEnrich(path_afterPC1_PC4$KEGG_2021_Human, showTerms = 10, numChar=60, y = "Ratio", orderBy = "Adjusted.P.value", title="KEGG_2021_Human") + scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) + scale_fill_viridis(begin=0.4, option="A")


# save pathway table
#write.xlsx(path_afterPC1_PC4, "/mnt/dzl_bioinf/calverom/Deconvolution/music_gse150910/supp_tables/pathwaytable_after_PC1_PC4correction.xlsx", rowNames=FALSE, colNames=FALSE)

```


# Correlation, all genes vs estimated cell proportions, PC1-PC4 correction
```{r}

corrRes_CellPropxallGene_afterPC1_PC4 <- NULL

for(ct in setdiff(celltypes,none)){
  corr <- apply(bulkNormLog2cpm, 1, cor.test, dfEst[,ct], method="pearson")
  genes <- names(corr)
  corrRes_CellPropxallGene_afterPC1_PC4 <- rbind(corrRes_CellPropxallGene_afterPC1_PC4, data.frame(cellType=ct, gene=genes, matrix(unlist(corr), nrow=length(corr), byrow=TRUE)))
}

corrRes_CellPropxallGene_afterPC1_PC4 <- corrRes_CellPropxallGene_afterPC1_PC4[,c(1:2,6,3,5)]
colnames(corrRes_CellPropxallGene_afterPC1_PC4)[3:5] <- c("corrCoefficient", "testStatistic", "pVal")
corrRes_CellPropxallGene_afterPC1_PC4[,3:5] <- lapply(corrRes_CellPropxallGene_afterPC1_PC4[,3:5], as.numeric)
corrRes_CellPropxallGene_afterPC1_PC4$pValDEA <- res_afterPC1_PC4$P.Value[match(corrRes_CellPropxallGene_afterPC1_PC4$gene, res_afterPC1_PC4$GeneName)]
corrRes_CellPropxallGene_afterPC1_PC4$pAdj <- p.adjust(corrRes_CellPropxallGene_afterPC1_PC4$pVal, method="BH")
corrRes_CellPropxallGene_afterPC1_PC4[order(corrRes_CellPropxallGene_afterPC1_PC4$pAdj),]
corrRes_CellPropxallGene_afterPC1_PC4$updown <- res_afterPC1_PC4$updown[match(corrRes_CellPropxallGene_afterPC1_PC4$gene, res_afterPC1_PC4$GeneName)]
# save corr table
#write.xlsx(corrRes_CellPropxallGene_afterPC1_PC4, "/mnt/dzl_bioinf/calverom/Deconvolution/music_gse150910/supp_tables/correlation_celltype_genes_afterPC1_PC4correction.xlsx", rowNames=FALSE, colNames=TRUE)

for(ct in orderCT){
  print(ggplot(corrRes_CellPropxallGene_afterPC1_PC4[which(corrRes_CellPropxallGene_afterPC1_PC4$cellType == ct),], aes(x=corrCoefficient, y=-log10(pValDEA))) + geom_point(aes(color=updown), alpha=0.4) + scale_color_manual(name="up/down reg", values=c("blue", "black", "red")) + xlab("correlation coefficient") + ylab("-log10(DEA p-value)") + annotate("text", x=-Inf, y=Inf, label=ct, hjust=-0.1, vjust=1, size=5) + theme_classic() + theme(legend.position="top"))
  pathh <- paste("./main_figures/ears_PC1PC4/ears_afterPC1PC4", "_", ct, ".jpg", sep="")
  #ggsave(filename=pathh, width=7, height=5, device="jpeg", dpi=350)
}


```

## Enrichr, cell type-specific pathways in bulk, PC1-PC4 correction
```{r}

# For EnrichR

# subset with correlation coefficient > 0.5
corrRes_afterPC1_PC4_sub <- corrRes_CellPropxallGene_afterPC1_PC4[corrRes_CellPropxallGene_afterPC1_PC4$corrCoefficient >= 0.5,]
corrRes_afterPC1_PC4_sub <- corrRes_afterPC1_PC4_sub[,-c(6,8)]
#corrRes_afterPC1_PC4_sub$gene2 <- geneAnno$GeneName[match(corrRes_afterPC1_PC4_sub$gene, geneAnno$ensembl)]
corrGenes <- unique(corrRes_afterPC1_PC4_sub$gene)
#corrGenes_conv <- geneAnno$GeneName[match(corrGenes, geneAnno$ensembl)]

# subset DE results with only genes from above^
res_afterPC1_PC4_sub <- res_afterPC1_PC4[res_afterPC1_PC4$GeneName %in% corrGenes,]
res_afterPC1_PC4_sub <- res_afterPC1_PC4_sub[,1:ncol(res_afterPC1_PC4_sub)-1]
# recompute adj.P.Val
res_afterPC1_PC4_sub$adj.P.Val <- p.adjust(res_afterPC1_PC4_sub$P.Value, method="BH")

# add p-values and adjusted p-values from above^ to corrRes_afterPC1_PC4_sub
corrRes_afterPC1_PC4_sub$pValDEA <- res_afterPC1_PC4_sub$P.Value[match(corrRes_afterPC1_PC4_sub$gene, res_afterPC1_PC4_sub$GeneName)]
corrRes_afterPC1_PC4_sub$pAdjDEA <- res_afterPC1_PC4_sub$adj.P.Val[match(corrRes_afterPC1_PC4_sub$gene, res_afterPC1_PC4_sub$GeneName)]

corrRes_afterPC1_PC4_sub$sig0.05 <- ifelse(corrRes_afterPC1_PC4_sub$pAdjDEA <= 0.05, "sig", "not sig")
corrRes_afterPC1_PC4_sub$sig0.1 <- ifelse(corrRes_afterPC1_PC4_sub$pAdjDEA <= 0.1, "sig", "not sig")
# count DE genes (alpha = 0.05) per cell type
s0.05 <- as.data.frame(table(corrRes_afterPC1_PC4_sub$cellType, corrRes_afterPC1_PC4_sub$sig0.05))
# count DE genes (alpha=0.1) per cell type
s0.1 <- as.data.frame(table(corrRes_afterPC1_PC4_sub$cellType, corrRes_afterPC1_PC4_sub$sig0.1))

# write.xlsx(corrRes_afterPC1_PC4_sub, "./results tree 22.4.25/afterPC1_PC4_correction_corr_CTxsubGenes.xlsx")

ggplot(s0.05[s0.05$Var2 == "sig",], aes(x=Var1, y=Freq, fill=Var2)) + geom_bar(stat="identity", position="dodge") + xlab("") + ylab("number of genes (for Enrichr)") + theme_classic() + coord_flip() + theme(legend.position="none")

```

```{r, out.extra='style="max-width:none; width:80vw; margin-left:calc(50% - 40vw);"', fig.width = 17, fig.height=10}

path_afterPC1_PC4_celltype_specific <- NULL
for(ct in orderCT){
  temp <- droplevels(corrRes_afterPC1_PC4_sub[corrRes_afterPC1_PC4_sub$cellType == ct, ])
  genes <- temp$gene[temp$sig0.05 == "sig"]
  if(length(genes) >= 10){
    enr <- enrichr(genes, dbs)
    path_afterPC1_PC4_celltype_specific[[ct]] <- enr
    p1 <- plotEnrich(path_afterPC1_PC4_celltype_specific[[ct]]$BioPlanet_2019, showTerms = 10, numChar=60, y = "Ratio", orderBy = "P.value", title="BioPlanet_2019") + scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) + scale_fill_viridis(begin=0.4, option="A") + ggtitle("")
    pathh <- paste("./main_figures/pathways_celltype_PC1PC4/path_afterPC1PC4", "_", ct, ".jpg", sep="")
    #ggsave(filename=pathh, plot=p1, width=7, height=5, device="jpeg", dpi=350)
    
    p2 <- plotEnrich(path_afterPC1_PC4_celltype_specific[[ct]]$WikiPathway_2021_Human, showTerms = 10, numChar=60, y = "Ratio", orderBy = "P.value", title="WikiPathway_2021_Human") + scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) + scale_fill_viridis(begin=0.4, option="A")
    p3 <- plotEnrich(path_afterPC1_PC4_celltype_specific[[ct]]$Reactome_2016, showTerms = 10, numChar=60, y = "Ratio", orderBy = "P.value", title="Reactome_2016") + scale_x_discrete(labels = function(x) str_wrap(x, width = 30)) + scale_fill_viridis(begin=0.4, option="A")
    title <- ggdraw() + draw_label(ct, x=0, hjust=0, size=12) + theme(plot.margin = margin(0, 0, 0, 7))
    pl <- plot_grid(p1,p2,p3, ncol=2)
    print(plot_grid(title,pl, ncol=1,rel_heights=c(0.1, 1)))
  }
}

pathways <- c("Extracellular matrix organization", "Wnt signaling pathway and pluripotency", "Wnt signaling pathway", "TGF-beta signaling pathway", "TGF-beta regulation of extracellular matrix", "TGF-beta signaling in development")
for(ct in names(path_afterPC1_PC4_celltype_specific)){
  past <- paste(ct, pathways[pathways %in% path_afterPC1_PC4_celltype_specific[[ct]][["BioPlanet_2019"]][["Term"]]])
  print(past)
}

# write to excel, pathway results per cell type
#write_rds(path_afterPC1_PC4_celltype_specific, "/mnt/dzl_bioinf/calverom/Deconvolution/music_gse150910/CTpathways_PC1_PC4corrected.rds")
#write_rds(path_afterPC1_PC4_celltype_specific, "/mnt/dzl_bioinf/calverom/Deconvolution/music_gse150910/supp_tables/CTpathways_PC1_PC4corrected.rds")

```

